---
title: "Tidyverse Create"
author: "Frederick Jones"
date: "2023-11-13"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This project is all about divorce rate in variaous socio-economic section in the US. The source for all data is Decennial Census (years 1960 to 2000) and American Community Survey (years 2001-2012), via IPUMS USA. 
The summary of variable names is as follows: 
<style>
table {
  font-family: arial, sans-serif;
  border-collapse: collapse;
  width: 100%;
}

td, th {
  border: 1px solid #dddddd;
  text-align: left;
  padding: 8px;
}

tr:nth-child(even) {
  background-color: #dddddd;
}
</style>
<table>
  <tr>
    <th>Header</th>
    <th>Description</th>
  </tr>
  <tr>
    <td>all</td>
    <td>Total (or all men/women in sex-specific files)</td>
  </tr>
  <tr>
    <td>HS</td>
    <td>High school graduate or less (EDUCD &lt; 65)</td>
  </tr>
  <tr>
    <td>SC</td>
    <td>Some college (EDUCD &gt;= 65 &amp; &lt;= 100)</td>
  </tr>
  <tr>
    <td>BAp</td>
    <td>Bachelor's degree or more (EDUCD &gt; 100)</td>
  </tr>
  <tr>
    <td>BAo</td>
    <td>Bachelor's degree, no graduate degree (EDUCD &gt; 100 &amp; &lt;= 113)</td>
  </tr>
  <tr>
    <td>GD</td>
    <td>Graduate degree (EDUCD &gt; 113)</td>
  </tr>
  <tr>
    <td>White</td>
    <td>Non-Hispanic white</td>
  </tr>
  <tr>
    <td>Black</td>
    <td>Black or African-American</td>
  </tr>
  <tr>
    <td>Hisp</td>
    <td>Hispanic of any race</td>
  </tr>
  <tr>
    <td>NE</td>
    <td>New England (REGION == 11)</td>
  </tr>
  <tr>
    <td>MA</td>
    <td>Mid-Atlantic (REGION == 12)</td>
  </tr>
  <tr>
    <td>Midwest</td>
    <td>Midwest (REGION == 21-23)</td>
  </tr>
  <tr>
    <td>South</td>
    <td>South (REGION == 31-34)</td>
  </tr>
  <tr>
    <td>Mountain</td>
    <td>Mountain West (REGION == 41)</td>
  </tr>
  <tr>
    <td>Pacific</td>
    <td>Pacific (REGION == 42)</td>
  </tr>
  <tr>
    <td>poor</td>
    <td>Family income in lowest 25%</td>
  </tr>
  <tr>
    <td>mid</td>
    <td>Family income in middle 50%</td>
  </tr>
  <tr>
    <td>rich</td>
    <td>Family income in top 25%</td>
  </tr>
  <tr>
    <td>work</td>
    <td>Employed 50+ weeks prior year</td>
  </tr>
  <tr>
    <td>nowork</td>
    <td>Not employed at least 50 weeks prior year</td>
  </tr>
  <tr>
    <td>nokids_all</td>
    <td>No own children living at home</td>
  </tr>
  <tr>
    <td>kids_all</td>
    <td>At least one own child living at home</td>
  </tr>
</table>


```{r}
df <- read.csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/marriage/divorce.csv')
head(df)
```
## Use of filter to filter to get the data for the variable 'all_3544'>0.1

```{r}
library(tidyverse)
library(ggplot2)
df<-df|> filter(
  all_3544>=0.1
)
df
```
## Use of mutate() to create a variable ratio = rich_4554/poor_4554

```{r}
df<- df|> mutate(
  ratio = rich_4554/poor_4554
)
df
```



